{\Large \bf Analysing time course microarray data using Bioconductor: a case study using yeast2 Affymetrix arrays}

### 2016-01-25 paper.Rnw
Typo: AffyPLM -> affyPLM.

The data were collected according to the experimental protocol described in
\cite{Greenall08}. Briefly, three biological replicates were studied on each of
a wild-type (WT) yeast strain and a strain carrying the \textit{cdc13-1}
temperature sensitive mutation (in which telomere uncapping is induced by growth
at temperatures above around 27${^\circ}$C). These replicates were sampled
initially at 23${^\circ}$C (at which \textit{cdc13-1} has essentially WT
telomeres) and then at 1, 2, 3 and 4 hours after a shift to 30${^\circ}$C to
induce telomere uncapping. The thirty resulting RNA samples were hybridised to
Affymetrix yeast2 arrays. The microarray data are available in the ArrayExpress
database (see \cite{Parkinson2009}) under accession number E-MEXP-1551 .

@@

<<eval=FALSE,tidy=FALSE>>=
##From Bioconductor
biocLite(c('ArrayExpress', 'Mfuzz, 'timecourse', 'yeast2.db', 
    'yeast2probe', 'yeast2cdf', aAffyPLM'))
##From cran
install.packages(c('GeneNet', 'gplots'))
@

@@

<<eval=TRUE,hide=TRUE,cache=TRUE,message=FALSE>>=
library(ArrayExpress)
yeast.raw = ArrayExpress('E-MEXP-1551')
@

@@

{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3163 kb)
    cdf = Yeast_2 (10928 affyids)
    number of samples = 30
    number of genes = 10928
    annotation = yeast2
\end{verbatim}}
\noindent If the Affymetrix microarray data sets have been downloaded into a single directory, then the \cc{.cel} files can be loaded into R using the \cc{ReadAffy} command.

Also available from ArrayExpress are the experimental conditions. However, some preprocessing is necessary:

<<eval=TRUE,cache=TRUE,tidy=FALSE>>=
ph = yeast.raw@phenoData
exp_fac = data.frame(data_order = 1:30, 
    strain = ph@data$Factor.Value..phenotype.,  
    replicates = ph@data$Characteristics..individual.,
    tps = ph@data$Factor.Value..time.)
levels(exp_fac$strain) = c('m', 'w')
exp_fac = with(exp_fac, exp_fac[order(strain, replicates, tps), ])
exp_fac$replicate = rep(1:3, each=5, 2)
@

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
stringsAsFactors=FALSE)
source('ExtractIDs.R')
c_df = ExtractIDs(s_cer[ ,1])
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library(affy)
library(yeast2probe)
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(s_cer[, 1], cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3167 kb)
    cdf = Yeast_2 (5900 affyids)
    number of samples = 30
    number of genes = 5900
    annotation = yeast2
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is 5,900 now that the \textit{S. pombe} probesets have been removed.

@@

<<s_fig3,eval=FALSE, echo=-1 >>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01,cex.axis=0.9, las=1, mfrow=c(1,2))
library(affyPLM)
#Raw data intensities
boxplot(yeast.raw, col='red', main='', ylim=c(2,16))
#Normalised intensities
boxplot(yeast.rma, col='blue', ylim=c(2,16))
@

@@

This package assesses treatment differences by comparing time-course
mean profiles allowing for variability both within and between time
points. It uses the multivariate empirical Bayes model proposed by
\cite{Tai06}. Further details of the \texttt{timecourse} package can be
found in \cite{Tai07}. After installing the \texttt{timecourse} package,
we construct a \texttt{size} matrix describing the replication
structure using
<<eval=TRUE,message=FALSE>>=
library(timecourse)
size = matrix(3, nrow = 5900, ncol = 2)
@
\noindent To extract a list of differentially expressed we calculate the
Hotelling statistic $\tilde T^2$ via
<<eval=TRUE,tidy=FALSE, cache=TRUE>>=
c.grp = as.character(exp_fac$strain)
t.grp = as.numeric(exp_fac$tps)
r.grp = as.character(exp_fac$replicate)
MB.2D = mb.long(yeast.matrix, times = 5, method = '2', 
    reps = size, condition.grp = c.grp, time.grp = t.grp, 
    rep.grp = r.grp)
@
\noindent The top (say) one hundred genes can be extracted via
<<eval=TRUE,cache=TRUE>>=
gene_positions = MB.2D$pos.HotellingT2[1:100]
gnames = rownames(yeast.matrix)
gene_probes = gnames[gene_positions]
@
\noindent The expression profiles can also be easily obtained. The profile for the top ranked expression is found using
<<s_fig4,eval=FALSE, echo=-1>>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1)
plotProfile(MB.2D, ranking=1, gnames=rownames(yeast.matrix))
@

@@

<<eval=TRUE,tidy=FALSE, message=FALSE>>=
library(limma)
expt_structure = factor(colnames(yeast.matrix))

@@

The following code (adapted from lecture material found at
\cite{speedcourse}) plots the time course expression for the top
one hundred differentially expressed genes according to their
$F$-statistic (see Figure~\ref{F2})

@@

<<F4,eval=TRUE,dev="pdf",fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Clustering of the top fifty differentially expressed genes. Red and green correspond to up- and down-regulation respectively.",message=FALSE, nice.par=TRUE,tidy=FALSE>>=
library(gplots)
#Cluster the top 50 genes
heatmap.2(m[1:50,], dendrogram ='row', Colv=FALSE, col=greenred(75), 
          key=FALSE, keysize=1.0, symkey=FALSE, density.info='none', 
          trace='none', colsep=rep(1:10), sepcolor='white', 
          sepwidth=0.05, labRow = NA, cexCol=1,
          hclustfun=function(c){hclust(c, method='average')})
@

@@

<<F5,eval=TRUE,dev="pdf",fig.ext='pdf',fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Eight clusters obtained from the \\texttt{Mfuzz} package.",message=FALSE>>=
library(Mfuzz)
tmp_expr = new('ExpressionSet', exprs=m)
cl = mfuzz(tmp_expr, c=8, m=1.25)
mfuzz.plot(tmp_expr,cl=cl, mfrow=c(2, 4), new.window = FALSE)
@
Of course, it is usually not clear how many clusters there are (or
should be) within a dataset and so the sensitivity of conclusions to
the choice of number of clusters ($c$) should always be
investigated. For example, if $c$ is chosen to be too large then some
clusters will appear sparse and this might suggest choosing a smaller
value of~$c$. Figure~\ref{F5} shows the profiles of the
eight clusters obtained from the \cc{Mfuzz} package. The probes 
present within each cluster can be found by using

@@

<<eval=TRUE,results='hide',message=FALSE>>=
exp_fac = with(exp_fac, exp_fac[order(strain, tps, replicates), ])
#Construct a longitudinal object
library(GeneNet)
ngenes = 100
m = yeast.matrix[ii[1:ngenes],]
mnew = m[,exp_fac$data_order[1:15]]
mlong = as.longitudinal(t(mnew), repeats=3, time=0:4)
@

@@

<<s_fig11,ref.label="s_fig1",eval=TRUE,dev="png",fig.ext="png",cache=FALSE,dpi=300,out.width="0.8\\linewidth",echo=FALSE, fig.cap="Image plots of the mismatch and perfect match probe intensities for the first replication of the mutant yeast strain. The corresponding times are indicated in the plot.", fig.pos="h">>=
#knitr::run_chunk("s_fig1")
@

>>>>

The data were collected according to the experimental protocol described in
\cite{Greenall08}. Briefly, three biological replicates were studied on each of
a wild-type (WT) yeast strain and a strain carrying the \textit{cdc13-1}
temperature sensitive mutation (in which telomere uncapping is induced by growth
at temperatures above around 27${^\circ}$C). These replicates were sampled
initially at 23${^\circ}$C (at which \textit{cdc13-1} has essentially WT
telomeres) and then at $1$, $2$, $3$ and $4$ hours after a shift to 30${^\circ}$C to
induce telomere uncapping. The thirty resulting RNA samples were hybridised to
Affymetrix yeast2 arrays. The microarray data are available in the ArrayExpress
database (see \cite{Parkinson2009}) under accession number E-MEXP-1551 .

@@

<<eval=FALSE,tidy=FALSE>>=
##From Bioconductor
biocLite(c('ArrayExpress', 'Mfuzz', 'timecourse', 'yeast2.db', 
    'yeast2probe', 'yeast2cdf', 'affyPLM'))
##From cran
install.packages(c('GeneNet', 'gplots'))
@

@@

<<eval=TRUE,hide=TRUE,cache=TRUE,message=FALSE>>=
library('ArrayExpress')
yeast.raw = ArrayExpress('E-MEXP-1551')
@

@@

{\small\begin{verbatim}
    AffyBatch object
    size of arrays=496x496 features (53 kb)
    cdf=Yeast_2 (5900 affyids)
    number of samples=30
    number of genes=5900
    annotation=yeast2
    notes=c("E-MEXP-1551", "", "", "", "", "")
	    c("E-MEXP-1551", "", "", "", "", "")
	    c("temperature", "phenotype", "time", "", "", "")
	    c("time_series_design", "genetic_modification_design", 
	        "stimulus_or_stress_design", "in_vitro_design", 
	        "co-expression_design", "transcription profiling by array")
\end{verbatim}}
\noindent If the Affymetrix microarray data sets have been downloaded into a single directory, then the \cc{.cel} files can be loaded into R using the \cc{ReadAffy} command.

Also available from \cc{ArrayExpress} are the experimental conditions. However some preprocessing is necessary:

<<eval=TRUE,cache=TRUE,tidy=FALSE>>=
ph = yeast.raw@phenoData
exp_fac = data.frame(data_order = 1:30, 
    strain = ph@data$Factor.Value..phenotype.,  
    replicates = ph@data$Characteristics..individual.,
    tps = ph@data$Factor.Value..time.)
levels(exp_fac$strain) = c('m', 'w')
exp_fac = with(exp_fac, exp_fac[order(strain, replicates, tps), ])
exp_fac$replicate = rep(1:3, each=5, times=2)
@

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
    stringsAsFactors=FALSE)
source('ExtractIDs.R')
c_df = ExtractIDs(s_cer[ ,1])
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library('affy')
library('yeast2probe')
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(s_cer[, 1], cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
   AffyBatch object
    size of arrays=496x496 features (53 kb)
    cdf=Yeast_2 (5900 affyids)
    number of samples=30
    number of genes=5900
    annotation=yeast2
    notes=c("E-MEXP-1551", "", "", "", "", "")
    	c("E-MEXP-1551", "", "", "", "", "")
    	c("temperature", "phenotype", "time", "", "", "")
    	c("time_series_design", "genetic_modification_design", 
    	"stimulus_or_stress_design", "in_vitro_design", 
    	"co-expression_design", "transcription profiling by array")
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is $5,900$ now that the \textit{S. pombe} probesets have been removed.

@@

<<s_fig3,eval=FALSE, echo=-1 >>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01,cex.axis=0.9, las=1, mfrow=c(1,2))
library('affyPLM')
#Raw data intensities
boxplot(yeast.raw, col='red', main='', ylim=c(2, 16))
#Normalised intensities
boxplot(yeast.rma, col='blue', ylim=c(2, 16))
@

@@

This package assesses treatment differences by comparing time-course
mean profiles allowing for variability both within and between time
points. It uses the multivariate empirical Bayes model proposed by
\cite{Tai06}. Further details of the \texttt{timecourse} package can be
found in \cite{Tai07}. After installing the \texttt{timecourse} package,
we construct a \texttt{size} matrix describing the replication
structure using
<<eval=TRUE,message=FALSE>>=
library('timecourse')
size = matrix(3, nrow = 5900, ncol = 2)
@
\noindent To extract a list of differentially expressed we calculate the
Hotelling statistic $\tilde T^2$ via
<<eval=TRUE,tidy=FALSE, cache=TRUE>>=
c.grp = as.character(exp_fac$strain)
t.grp = as.numeric(exp_fac$tps)
r.grp = as.character(exp_fac$replicate)
MB.2D = mb.long(yeast.matrix, times = 5, method = '2', 
    reps = size, condition.grp = c.grp, time.grp = t.grp, 
    rep.grp = r.grp)
@
\noindent The top (say) one hundred genes can be extracted via
<<eval=TRUE,cache=TRUE>>=
gene_positions = MB.2D$pos.HotellingT2[1:100]
gnames = rownames(yeast.matrix)
gene_probes = gnames[gene_positions]
@
\noindent The expression profiles can also be easily obtained. The profile for the top ranked expression is found using
<<s_fig4,eval=FALSE, echo=-1>>=
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1)
plotProfile(MB.2D, ranking=1, gnames=rownames(yeast.matrix))
@

@@

<<eval=TRUE,tidy=FALSE, message=FALSE>>=
library('limma')
expt_structure = factor(colnames(yeast.matrix))

@@

\noindent The following code (adapted from lecture material found at
\cite{speedcourse}) plots the time course expression for the top
one hundred differentially expressed genes according to their
$F$-statistic (see Figure~\ref{F2})

@@

<<F4,eval=TRUE,dev="pdf",fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Clustering of the top fifty differentially expressed genes. Red and green correspond to up- and down-regulation respectively.",message=FALSE, nice.par=TRUE,tidy=FALSE>>=
library('gplots')
# Cluster the top 50 genes
heatmap.2(m[1:50,], dendrogram ='row', Colv=FALSE, col=greenred(75), 
          key=FALSE, keysize=1.0, symkey=FALSE, density.info='none', 
          trace='none', colsep=rep(1:10), sepcolor='white', 
          sepwidth=0.05, labRow = NA, cexCol=1,
          hclustfun=function(c){hclust(c, method='average')})
@

@@

<<F5,eval=TRUE,dev="pdf",fig.ext='pdf',fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Eight clusters obtained from the \\texttt{Mfuzz} package.",message=FALSE>>=
library('Mfuzz')
tmp_expr = new('ExpressionSet', exprs=m)
cl = mfuzz(tmp_expr, c=8, m=1.25)
mfuzz.plot(tmp_expr,cl=cl, mfrow=c(2, 4), new.window = FALSE)
@
Of course, it is usually not clear how many clusters there are (or
should be) within a dataset and so the sensitivity of conclusions to
the choice of number of clusters ($c$) should always be
investigated. For example, if $c$ is chosen to be too large then some
clusters will appear sparse and this might suggest choosing a smaller
value of~$c$. Figure~\ref{F5} shows the profiles of the
eight clusters obtained from the \cc{Mfuzz} package. The probes 
present within each cluster can be found by using

@@

<<eval=TRUE,results='hide',message=FALSE>>=
exp_fac = with(exp_fac, exp_fac[order(strain, tps, replicates), ])
#Construct a longitudinal object
library('GeneNet')
ngenes = 100
m = yeast.matrix[ii[1:ngenes],]
mnew = m[,exp_fac$data_order[1:15]]
mlong = as.longitudinal(t(mnew), repeats=3, time=0:4)
@

@@

<<F1A,ref.label="s_fig1",eval=TRUE,dev="png",fig.ext="png",cache=FALSE,dpi=300,out.width="0.8\\linewidth",echo=FALSE, fig.cap="Image plots of the mismatch and perfect match probe intensities for the first replication of the mutant yeast strain. The corresponding times are indicated in the plot.", fig.pos="h">>=
#knitr::run_chunk("s_fig1")
@
